---
title: "Jamaica again"
author: "Enrique Delamonica"
date: "2026-10-04"
output: github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```


## Introduction

In this exercise, I try to redo some of the steps I was not able to accomplish last week. Thank you for the opportunity to do so!

As usual, first, some libraries and the data are downloaded.

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
```

```{r}
url<-"https://raw.githubusercontent.com/Enrique01234/607-Fall-2026/refs/heads/main/Jamaica%20BReg%20again%20shorter.csv"
```

## Good news

The raw data from GitHub are converted into a data frame. 

```{r}
JamBReg <- read_csv(file = url, show_col_types = FALSE, progress = FALSE)
```

```{r}
JamBReg
```

Many of my problems were fixed just by using the "raw" button in GitHub!

This looks like quite a tidy data frame. Nevertheless, I tried to clean up the table a bit. First, I eliminated the lines which are separator rows. I created a new data frame (JamBReg2). As they contain NAs, I used (which is a simpler task than what Jocelyn did in the ELO case):

```{r}
JamBReg2 <- JamBReg|>
  drop_na()
```


```{r}
JamBReg2
```

Then, I tried to replace the missing values with NAs. However, I could not do so. It would have to be done in two steps, one for "(*)" and another one for "-". They do have a different meaning. The latter means there were no observations while the former means there were too few observations and a reliable value cannot be reported. For purposes of calculations, neither can be used so both can be labeled as missing values.


## Next steps

First, the lines that I have  dropped do have useful information which I will need later on because they are the axes of disparities I would have to find a way to add them. One problem is that the number of categories in each group is different. Neverthless, that information needs to be included.

Secondly, I need to replace the missing values.
